Score-based tests of measurement invariance: Use in practice

### 2013-09-17 frontiers/frontiers.Rnw
update refenrence and general discussion section.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@260 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

In this section, we demonstrate how the above tests can be carried out in~R, 
using the package lavaan \cite{lavaan11} for model estimation and strucchange 
\cite{ZeiLei02,Zei06} for testing.  We use data from \citeA{FroFan11} concerning 
the applicability of adult gratitude scales to youth, available in the~R 
package psychotools \cite{zeilei11}.  The data consist of responses to three adult 
gratitude scales from $n=1401$ youth aged 10--19 years. The original authors 
were specifically interested in whether the scales were measurement invariant 
across age.  Because the sample size at each age was unbalanced, the authors 
created age groups of approximately equal sample size.  In the examples below, 
we test for measurement invariance across these age groups.

@@

<<ordtest,eval=FALSE>>=
wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="WDMo")
maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLMo")

@ 
<<pvals>>=
round(c(wdmo$p.value, maxlmo$p.value), 2)
@

@@

Power and Type I error were examined across three sample sizes
($n=120,480,960$), three numbers of categories ($m=4,8,12$) and
17 magnitudes of invariance violations. 
The measurement invariance violations began at level $1+m/2$ of $V$
and were consistent thereafter. These
violations were based on the auxiliary variable $V$: individuals below
the $1+m/2$ percentile of $V$ deviated from individuals above the $1+m/2$ 
percentile  by $d$ times the parameters' asymptotic standard errors
(scaled by $\sqrt n$), with $d=0,0,25,0.5,...,4$. 
For each combination of sample size ($n$) $\times$ violation
magnitude ($d$) $\times$ violating parameter $\times$ categories $m$, 5000 datasets were
generated and tested. Four statistics were examined above. In all conditions, we
maintained equal sample sizes at each level of the ordinal variable.

@@

\section{Simulation 2}
In Simulation 2, we examine the extent to which the results of
Simulation 1 are robust to model misspecification. Specifically, we
generate data from the factor analysis model used in the previous
section, but with adding an extra loading from the second factor to the first
indicator (of the first factor).  The estimated model lacks this extra
loading, however. The goal of this simulation is to examine the proposed statistics'
power and specificity under model misspecification.

@@

In this paper, we examined the recently proposed ordinal statistics specificity property and robustness to model specification to study measurement invariance. As demonstrated in the results, the proposed statistics could select the parameter violating measurement invariance even though the model is misspecified in general. However, when an unmodeled parameter is violating measurement invariance, the statistics will demonstrating these variances in the related parameters, for example, associated loading and error terms.

These ordinal tests are useful in behavioral and social science in general, since the truly continuous auxiliary variable is hard to find in real research context. It is very common to observe several individuals with the same age, IQ, income rank, etc. Therefore, the ordinal tests which dealing with multiple ties in auxiliary variable is more suitable than the continuous test.

@@

We focused on testing for measurement invariance in factor analysis model in this paper, however, the core of the proposed ordinal test is to get the cumulative score process, which can be applied to any ML estimation functions or functions relating to maximizing or minimizing some statistics quantity. For example, studying DIF in IRT is immediate.

In summary, the ordinal test statistic proposed have relatively-high power for detecting measurement invariance violation parameter specifically that are monotonic with the auxiliary ordinal variable. This
feature is robust as long as the violation parameter is included in the testing model. Furthermore, as illustrated in tutorial part, the use of these tests is convenient without changing the focal psychometric model. In all, the tests have advantageous properties that should be implemented in practice.

>>>>

In this section, we demonstrate how the above tests can be carried out in~R, 
using the package lavaan \cite{lavaan11} for model estimation and strucchange 
\cite{ZeiLei02,Zei06} for testing.  We use data from \citeA{FroFan11} concerning 
the applicability of adult gratitude scales to youth, available in the~R 
package psychotools \cite{zeilei12}.  The data consist of responses to three adult 
gratitude scales from $n=1401$ youth aged 10--19 years. The original authors 
were specifically interested in whether the scales were measurement invariant 
across age.  Because the sample size at each age was unbalanced, the authors 
created age groups of approximately equal sample size.  In the examples below, 
we test for measurement invariance across these age groups.

@@

<<ordtest,eval=FALSE>>=
wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="WDMo")
maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLMo")
@ 
<<pvals>>=
round(c(wdmo$p.value, maxlmo$p.value), 2)
@

\seeme{I don't know how to adjust the line interval space for the code part.}

@@

\seeme{When I make the picture bigger, the symbols became blur. I think I might need to work on the plot tex file.}
Power and Type I error were examined across three sample sizes
($n=120,480,960$), three numbers of categories ($m=4,8,12$) and
17 magnitudes of invariance violations. 
The measurement invariance violations began at level $1+m/2$ of $V$
and were consistent thereafter. These
violations were based on the auxiliary variable $V$: individuals below
the $1+m/2$ percentile of $V$ deviated from individuals above the $1+m/2$ 
percentile  by $d$ times the parameters' asymptotic standard errors
(scaled by $\sqrt n$), with $d=0,0,25,0.5,...,4$. 
For each combination of sample size ($n$) $\times$ violation
magnitude ($d$) $\times$ violating parameter $\times$ categories $m$, 5000 datasets were
generated and tested. Four statistics were examined above. In all conditions, we
maintained equal sample sizes at each level of the ordinal variable.

@@

\seeme{I think it might be nicer to keep these plots all have three rows?(Not sure)}
\section{Simulation 2}
In Simulation 2, we examine the extent to which the results of
Simulation 1 are robust to model misspecification. Specifically, we
generate data from the factor analysis model used in the previous
section, but with adding an extra loading from the second factor to the first
indicator (of the first factor).  The estimated model lacks this extra
loading, however. The goal of this simulation is to examine the proposed statistics'
power and specificity under model misspecification.

@@

In this paper, we examined the recently proposed ordinal statistics specificity property and robustness to model specification to study measurement invariance. As demonstrated in the results, the proposed statistics could select the modeled parameter violating measurement invariance even though the model is misspecified in general.

\subsection{Application}
These ordinal tests are useful in behavioral and social science in general, since the truly continuous auxiliary variable is hard to find in real research context. It is very common to observe several individuals with the same age, IQ, income rank, etc. Therefore, the ordinal tests which dealing with multiple ties in auxiliary variable is more suitable than the continuous test.

@@

\subsection{Extension}
We focused on testing for measurement invariance in factor analysis model in this paper, however, the core of the proposed ordinal test is to get the cumulative score process, which can be applied to any ML estimation functions or functions relating to maximizing or minimizing some statistics quantity. For example, studying DIF in IRT is immediate.

\subsection{Conclusion}
In summary, the ordinal test statistic proposed have relatively-high power for detecting measurement invariance violation parameter specifically that are monotonic with the auxiliary ordinal variable. This
feature is robust as long as the violation parameter is included in the testing model. Furthermore, as illustrated in tutorial part, the use of these tests is convenient without changing the focal psychometric model. In all, the tests have advantageous properties that should be implemented in practice.
